---
title: "Bayesian Modeling and Simulation: Day 2"
date: 2025-08-19
---

# Questions/review

- ?

# Today's agenda

- Installing all the stuff
- Stan basics:
  - creating a .stan file
  - running a Stan model in R
- MCMC basics
  - chains, iter
  - inits
  - warm-up/burn-in
- Examining posterior samples
  - quantiles
  - HPDI
- Playing with our simple regression
  - play with init, warmup
  - play with priors

# Installing all the stuff

- RTools if you're on windows:

https://cran.r-project.org/bin/windows/Rtools/

LOTS of packages out there. I actually recommend loading as few as needed at a time. Many packages overlap in functionality and functions, and you can sometimes get unexpected things happening.

Core packages for running Stan:

```{r, eval = FALSE}
install.packages(c("rstan", "rstanarm"))
```

Handy additional packages we'll use:

```{r, eval = FALSE}
install.packages(c("bayesplot", "brms", "HDIntverval"))
```

For the "Statistical Rethinking" textbook (not on CRAN):

```{r, eval = FALSE}
devtools::install_github("rmcelreath/rethinking")
```

## Loading for today

Today we will just load the following:

```{r}
library(rstan)
library(rstanarm)
library(ggplot2)
library(bayesplot)
library(HDInterval)
library(HistData)
```


# Stan basics
  - creating a .stan file
  - running a Stan model in R

# MCMC basics
  - chains, iter
  - inits
  - warm-up/burn-in

# Examining posterior samples
  - quantiles
  - HPDI

# Playing with our simple regression
  - play with init, warmup
  - play with priors


